Why do we Eval?\\[2mm]\Large A large-scale study of Eval usage in R

### 2020-11-30 inst/paper/main.tex
background

\begin{lstlisting}
 eval
   <- function(expr,
                envir = parent.frame(),
                enclos = if(is.list(envir)) parent.frame() else baseenv())
          { ...body... }
\end{lstlisting}
The parameters are \c{expr}, the value to be evaluated, \c{envir}, the
environment in which evaluation happens, and \c{enclos}, the
environment to look up objects not found in \c{envir} (by default
either the top level, \c{baseenv} or the environment of the caller).
The \c{expr} parameter can take values of different type, for our
purposes we focus on the most common, {\tt expression}s. It is easiest
to think of an \c{expression} as an abstract syntax tree as returned
by R's parser. The simplest way to create an \c{expression} is to call
\c{quote} passing some R expression:
\begin{lstlisting}
  > quote(a + b)
  # a + b
\end{lstlisting}
Here the return value is an abstract syntax tree. From a
string, use  the \c{parse} function:
\begin{lstlisting}
  > parse(text="a+b")
  # expression(a+b)
\end{lstlisting}

There are four reflective functions
that take abstract syntax trees (called expressions), parse them as
code, and evaluate the result. These functions are \c{eval},
\c{evalq}, \c{eval.parent}, and \c{local}. Taking an expression rather
than a string, may discourage some casual users as manipulating
expressions is slightly more involved.\footnote{The rationale for
\eval taking expressions is explained by the main use case for \eval,
namely extending the language by reinterpretation of promises; R gives
access to the code of a promise as an expression.} There are two ways
to obtain an expression. Given a string, the \c{parse()} function
retuns an unevaluated expression. The other way is to use the function
\c{substitute(e,envir)}, it yields the parse tree of the expression
\c{e} after performing substitutions defined by the bindings in the
\c{envir} parameter.

\begin{lstlisting}
 > substitute(expression(a+b), list(a=1)))
 expression(1+b)
\end{lstlisting}

\noindent
This function is widely used in conjunction with \eval, as it can
leverage R's laziness for metaprogramming. R allows programmatic
manipulation of parse trees, which are themselves first class objects.
Expressions are evaluated using \eval (or one its variants). The \eval
function takes three arguments:

The \c{evalq} form is equivalent to \c{eval(quote(expr), ...)}.
Quoting that argument prevents if from being evaluated in the current
environment. \c{eval.parent(expr, n)} is a shorthand for \c{eval(expr,
  parent.frame(n))}. \c{local} evaluates an expression in a local
environment. It is equivalent to evalq except that its default
argument creates a new, empty environment. This is useful to create
anonymous recursive functions and as a kind of limited namespace
feature since variables defined in the environment are hidden.

\medskip\noindent\textit{Environments.} In R, environments can be
manipulated at will. Each function has an environment that can be
accessed with \c{environment()}. A new environment, which by default
inherits from the current environment, is created with \c{new.env}.
Contrary to other values, environments have reference semantics rather
than copy semantics. The chain of inheriting environments can be
traversed with \c{parent.env}, ultimately one arrives at \c{emptyenv}.
\c{parent.frame} and \c{sys.frame} grant access to environments
further up in the calling stack, up to the global environment
\c{.GlobalEnv}, and packages. One can also directly read, modify or
create new bindings in any environments. Given an environment \c{e}:

\begin{itemize}
\item  \c{e$v} or \c{get("v", envir = e)} reads variable \c{v} in \c{e};
\item \c{e$v <- 2} or \c{assign("v", 2, envir = e)} writes value 2 in variable \c{v} in \c{e}. \c{v} is created if it did not exist in \c{e}.
\end{itemize}
\noindent One can directly modify a variable in the scope of its
calling function, and there is even a special assignment operator for
that purpose; \c{v <<- 2} will search for \c{v} in parent
environments, redefine the value, or create a definition in the
current environment. Environments are used as hashtables, to maintain
state across function calls, or to create package namespaces.

>>>>

\begin{lstlisting}
 eval
   <- function(expr,
                envir = parent.frame(),
                enclos = if(is.list(envir)) parent.frame() else baseenv())
          { ...body... }
\end{lstlisting}
The parameters are \c{expr}, the value to be evaluated, \c{envir}, the
environment in which evaluation happens, and \c{enclos}, the
environment to look up objects not found in \c{envir} (by default
either the top level, \c{baseenv} or the environment of the caller).
The \c{expr} parameter can take values of different type, for our
purposes we focus on the most common, {\tt expression}s. It is easiest
to think of an \c{expression} as an abstract syntax tree as returned
by R's parser. The simplest way to create an \c{expression} is to call
\c{quote} passing some R expression:
\begin{lstlisting}
  > quote(a + b)
  # a + b
\end{lstlisting}
Here the return value is an abstract syntax tree. From a
string, use  the \c{parse} function:
\begin{lstlisting}
  > parse(text="a+b")
  # expression(a+b)
\end{lstlisting}
The most common way to create an expression is to extract it from an
argument. Each argument comes packaged inside a promise which retains
the source code of the argument. Consider the following function
definition, \c{f} is a function with a single parameter \c{x}. When
the call \c{f(a+b)} is evaluated, \c{a+b} is stored inside a newly
created promise.
\begin{lstlisting}
  > f <- function(x) substitute(x), list(a=1)))
  > f(a+b)
  # expression(1+b)
\end{lstlisting}
The call to \c{substitute(x)} extracts, from the promise bound to
\c{x}, the expression corresponding to that argument, i.e., \c{a+b}.
The general form of the function takes two arguments, the second
argument is an object that can be used as an environment. This can be
either a list of named values, a data frame or an actual environment.
\begin{lstlisting}
  > environment()
  # <environment: R_GlobalEnv>
\end{lstlisting}
The above shows how to access the current environment, this can belong
to the current function or, as here, the top level. Environments are
nested, with each environment having a parent. This nesting is used
when looking up names. When calling \c{new.env}, the default parent is
the current environment. The chain of environments can be traversed
with \c{parent.env}, ultimately one arrives at \c{emptyenv}.
\c{parent.frame} and \c{sys.frame} grant access to environments
further up in the calling stack, up to the global environment
\c{.GlobalEnv}, and packages. One can also directly read, modify or
create new bindings, given any environment:
\begin{itemize}
\item \c{env$v} and \c{get("v", envir = env)} read variable \c{v} in
  \c{e};
\item \c{env$v <- 2} and \c{assign("v", 2, envir = env)} write 2 in
  \c{v}. If it did not exist, \c{v} is created in \c{env}.
\end{itemize}
\noindent Environments are often used as hashtables by programmers as
they have reference semantics (other are copy-on-write) and have a
built-in string look up. across function calls, or to create package
namespaces.
\begin{lstlisting}
  evalq <- function(exp,env,enclos) eval(quote(exp),env,enclos)
  eval.parent <- function(exp,n) eval(exp,parent.frame(n))
  local <- function(exp,env,enclos) evalq(exp,new.env())
\end{lstlisting}
The above code show simplified implementation for the variants of
\eval. The \c{evalq} form simply quotes the argument to prevent it
from being evaluated in the current environment. The
\c{eval.parent(e,n)} form evaluates \c{e} in the environment of the
  {\tt n}-th caller up the call stack. Finally \c{local} evaluates an
  expression in a new environment to avoid polluting the current
  environment.

%% Taking an expression rather
%% than a string, may discourage some casual users as manipulating
%% expressions is slightly more involved.\footnote{The rationale for
%% \eval taking expressions is explained by the main use case for \eval,
%% namely extending the language by reinterpretation of promises; R gives
%% access to the code of a promise as an expression.}
